Formalization of Network Flow Algorithms

### 2017-03-02 paper/root.tex

\title{Formalization of Network Flow Algorithms}
\subtitle{a refinement approach in Isabelle/HOL}

@@

Using stepwise refinement techniques, we instantiate the generic Ford-Fulkerson method to Edmonds-Karp algorithm, 
and the generic push-relabel algorithm of Goldberg and Tarjan to both, the relabel-to-front and the FIFO push-relabel algorithm.
Further refinement then yields verified efficient implementations of the algorithms, which compare well to unverified reference implementations.

@@

\section{Introduction}
Computing the maximum flow of a network is an important problem in graph theory.
Many other problems, like maximum-bipartite-matching, edge-disjoint-paths, circulation-demand, as well as various scheduling and resource allocating problems can be reduced to it.
The Ford-Fulkerson method~\cite{FF56} describes a class of algorithms to solve the maximum flow problem. 
The basic idea is to repeatedly search for \emph{augmenting path} along which more flow can be moved from the source to the sink.
An important instance is the Edmonds-Karp algorithm~\cite{EK72},
which was one of the first algorithms to solve the maximum flow problem in polynomial time ($O(VE^2)$) for the general case of networks with real-valued capacities.
Dinitz~\cite{Di06} generalized the idea of augmenting paths to blocking flows, obtaining an $O(V^2E)$ algorithm. 
Karzanov~\cite{Ka74} was the first to propose an $O(V^3)$ algorithm based on the idea of \emph{preflows}. Based on Karzanov's ideas, Goldberg and Tarjan proposed
the generic push-relabel algorithm~\cite{GoTa88}.
Implementations of the push-relabel algorithm are among the most efficient maximum flow algorithms. While the generic algorithm has a time complexity of $O(V^2E)$,
specific variants of the algorithms achieve even lower complexities up to $O(V^2\sqrt{E})$.

In this paper, we present a formal verification of the Edmonds-Karp algorithm and of the generic push-relabel algorithm.
We prove correctness of the algorithms as well as the time complexity bounds of $O(VE^2)$ and $O(V^2E)$, respectively. 
The formalization is conducted in the Isabelle/HOL proof assistant~\cite{NPW02}. 
Stepwise refinement techniques~\cite{Wirth71,Back78,BaWr98} allow us to elegantly structure our verification to separate the abstract algorithmic ideas from
the implementation details: After proving the min-cut-max-flow theorem, it is straightforward to verify a generic version of the Ford-Fulkerson algorithm,
which we then refine to Edmonds-Karp algorithm and finally to an efficient implementation. 
Similarly, the generic push-relabel algorithm is refined to both, the relabel-to-front algorithm~\cite{CLRS09} and the FIFO push-relabel algorithm~\cite{GoTa88},
which, in turn, are refined to efficient implementations. The efficiency of our verified implementations compare well to unverified reference implementations of 
the algorithms.

@@

In the rest of this paper, we give a short informal introduction to flow networks and the Ford-Fulkerson method (\S\ref{sec:background}).
We then present our formalization of these material (\S\ref{sec:abs-formalization}). After introducing the Isabelle Refinement Framework~\cite{LaTu12,La12} (\S\ref{sec:refinement}), we report on the instantiation of the Ford-Fulkerson method to the Edmonds-Karp algorithm, and on the complexity proof (\S\ref{sec:edka}).
Next, we describe the correctness and complexity proof of the generic push-relabel algorithm (\S\ref{sec:gen-prpu}), and its instantiations 
to the relabel-to-front and FIFO push-relabel algorithms (\S\ref{sec:prpu-inst}).
We then report on the refinement of all three algorithms to efficient implementations (\S\ref{sec:executable}), and their comparison to unverified reference 
implementations (\S\ref{sec:benchmark}).

@@

A (flow) network is a directed graph over a finite set of vertices $V$ and edges $E$, where each edge $(u,v)\in E$ is labeled by a positive real-valued capacity $c(u,v)>0$.
Moreover, there are two distinct vertices $s,t\in V$, which are called \emph{source} and \emph{sink}.

A \emph{flow} $f$ on a network is a labeling of the edges with real values satisfying the following constraints: 1) \emph{Capacity constraint}: the flow on each edge is a non-negative value smaller or equal to the edge's capacity; 2) \emph{Conservation constraint}: For all vertices except $s$ and $t$, the sum of incoming flows equals the 
sum of outgoing flows.
% 
% \emph{Non-generation constraint}: For all vertices except $s$ and $t$, the sum of flows over all incoming edges is not smaller than the sum of flows over all outgoing edges.
% We define the \emph{excess} flow $x_f(u)$ of a node $u$ to be the difference of the incoming and outgoing flow of $u$.
% Intuitively, a \emph{preflow} must not send more flow over an edge than its capacity allows, and only the source node may generate flow. 
% The excess flow is the amount of flow that is stuck at a node, \ie which flows in but does not flow out.
% A \emph{flow} is a \emph{preflow} that satisfies the \emph{Conservation constraint}: For all vertices except $s$ and $t$, the excess flow is zero.
% Intuitively, only the source may generate flow, and only the sink may consume flow, and for all other nodes, the incoming flow must be equal to the outgoing flow.
The value of a flow $f$ is denoted by $|f|$, and defined to be the sum over the outgoing flows of $s$ minus the sum over the incoming flows of $s$.
Given a network $G$, the maximum flow problem is to find a flow with a maximum value among all flows of the network.

To simplify reasoning about the maximum flow problem, we assume that our network satisfies some additional constraints: 1) the source only has outgoing edges while the sink only has incoming edges; 2) if the network contains an edge $(u, v)$ then there is no \emph{parallel edge} $(v, u)$ in the reverse direction\footnote{With $u=v$, this also implies that there are no self loops.}; and 3) every vertex of the network must be on a path from $s$ to $t$. Note that any network can be transformed to a network with the aforementioned properties and the same maximum flow~\cite{CLRS09}.

An important result is the relation between flows and cuts in a network. A \emph{cut} is a partitioning of the vertices into two sets, such that one set contains 
the source and the other set contains the sink. The capacity of a cut is the sum of the capacities of all edges going from the source's side to the sink's side of the cut.
It is easy to see that the value of any flow cannot exceed the capacity of any cut, as all flow from the source must ultimately reach the sink, and thus go through the edges of the cut. The Ford-Fulkerson theorem tightens this bound and states that the value of the maximum flow is equal to the capacity of the minimum cut.

The Ford-Fulkerson method is a corollary of this theorem. It is based on a greedy approach: Starting from a zero flow, the value of the flow is iteratively increased until a maximum flow is reached. In order to increase the overall flow value, it may be necessary to redirect some flow, \ie to decrease the flow passed through specific edges. For this purpose the Ford-Fulkerson method defines the residual graph, which has edges in the same and opposite direction as the network edges.
% , which has edges both in parallel and reverse to the edges of the network. 
Each edge is labeled by the amount of flow that can be effectively passed along this edge, by either increasing or decreasing the flow on a network edge.
Formally, the residual graph $G_f$ of a (pre)flow $f$ is the graph induced by the 
edges with positive labels according to the following labeling function $c_f$:
\[ c_f (u, v) = 
  \begin{cases}
  c (u, v) - f(u, v) \hfill & \text{ if $(u, v) \in E$} \\
  f (v, u) \hfill & \text{ if $(v, u) \in E$} \\  
  0 \hfill & \text{ otherwise} \\
  \end{cases} 
\]

@@

The correctness of the Ford-Fulkerson algorithm follows from the Ford-Fulkerson theorem, which is also known as min-cut-max-flow theorem.
It states that the following three propositions are equivalent:
\begin{enumerate}
\item $f$ is a maximum flow in a network $G$.
\item there is no augmenting path in the residual graph $G_f$.
\item there is a cut $C$ in $G$ such that the capacity of $C$ is equal to the value of $f$.
\end{enumerate}

@@

In Isabelle/HOL, stepwise refinement is supported by the Isabelle Refinement Framework~\cite{LaTu12,La12}. 
It features a refinement calculus for programs phrased in a nondeterminism monad. 
The monad's type is a set of possible results plus an additional value that indicates a failure:
\begin{lstlisting}
  datatype \<alpha> nres = res \<alpha> set | fail
\end{lstlisting}
The operation \isai{return x} of the monad describes the single result \isai{x}, and the 
operation \isai{bind m f} nondeterministically picks a result from \isai{m} and executes \isai{f} on it. 
The bind operation fails iff either \isai{m = fail}, or \isai{f} may fail for a result in \isai{m},

We define the \emph{refinement ordering} on \isai{\<alpha> nres} by lifting the subset ordering with \isai{fail} being the greatest element.
Intuitively, \isai{m <= m'} means that \isai{m} is a refinement of \isai{m'}, \ie all possible results of \isai{m} are 
also possible results of \isai{m'}. 
Note that the refinement ordering is a complete lattice, and bind is monotonic. Thus, we can define recursion using a fixed-point construction~\cite{Kr10}.
Moreover, we can use the standard Isabelle/HOL constructs for if, let and case distinctions, yielding a fully fledged programming 
language, shallowly embedded into Isabelle/HOL's logic. For simpler usability, we define standard loop constructs (while, foreach), 
a syntax for postcondition specifications, and use a Haskell-like do-notation:
\begin{lstlisting}
  spec P == spec x. P x == res {x. P x}
  do {x <- m; f x} == bind m f
  do {m; m'} == bind m (%_. m')
\end{lstlisting}

Correctness of a program \isai{m} with precondition \isai{P} and postcondition \isai{Q} is expressed as \isai{P ==> m <= spec r. Q r} (or, eta-contracted, just \isai{spec Q}), which
means that, if \isai{P} holds, \isai{m} does not fail and all possible results of \isai{m} satisfy \isai{Q}. Note that we provide different recursion constructs
for partial and total correctness: A nonterminating total correct recursion yields \isai{fail}, which satisfies no specification, even if joined 
with results from other possible runs. On the other hand, a nonterminating partial correct recursion yields \isai|res {}|, which refines any specification 
and disappears when joined with other results.

@@

Various tools, including a verification condition generator, assist the user in conducting the refinement proofs by breaking 
them down to statements that do not contain monad constructs any more. In many cases, these verification conditions reflect the core idea of the proof precisely.

@@

\section{The Generic Push-Relabel Algorithm}\label{sec:gen-prpu}
  We present our formalization of the generic push relabel algorithm of Goldberg and Tarjan~\cite{GoTa88}. 
  Again, our formalization closely follows the textbook presentation of Cormen et al.~\cite{CLRS09}.

  Push-relabel algorithms are based on a generalization of flows called \emph{preflows}: 
  The conservation constraint is relaxed to allow more incoming flow than outgoing flow on a vertex.
  For a vertex $u\in V$, the \emph{excess} $x_f(u)$ is the difference between incoming and outgoing flow. 
  For a preflow, the excess of any node but the source must not be negative.
  A non-sink vertex with positive excess is called \emph{active}. Thus, a preflow without any active vertices is actually a flow.

  The push-relabel algorithm maintains a preflow $f$ and a height-labeling $l: V \to \mathbb {N}$ of the nodes, 
  with the invariant that for every edge $(u, v)$ of the residual graph we have that $l(u) \le l(v) + 1$. 
  Moreover, the height of the source is fixed to $\left\vert{V}\right\vert$, and the height of the sink is fixed to $0$.
  By generalizing the labeling invariant to paths, we get that for each path $p$ from $u$ to $v$ in the residual graph, we have $l(u) \le l(v) + |p|$.
  This implies that there cannot be a path from $s$ to $t$: The length of any (simple) path is less than $|V|$, however, a valid labeling 
  implies a length greater than or equal to $|V|$. Upon termination of the algorithm, there will be no active nodes. 
  Thus, the preflow is actually a flow, and as we have no $s$--$t$ paths, the Ford-Fulkerson theorem implies that it is maximal.

  The push relabel algorithm initializes the heights of all nodes but the source to zero. Moreover, the outgoing edges of the source are 
  set to carry maximum flow, all other edges get zero flow. Then, the algorithm repeats the following operation as long as they apply:
  \begin{description}
    \item[$\textrm{push}(u,v)$] for an edge $(u,v)$ of the residual graph with $l(u) = l(v)+1$ and $u$ active: Move as much flow as possible from $u$ to $v$.
    The maximum possible flow is constrained by the residual capacity of the edge, and the excess flow available at $u$.
    \item[$\textrm{relabel}(u)$] for an active node $u$ from which no push operation is possible: Increase the height of $u$ to the minimum value such that a push
      becomes possible.
  \end{description}
  Obviously, if none of the operations apply, all nodes are inactive and the resulting preflow is a maximum flow.
  The algorithm is generic in the sense that it leaves open the order in which the operations are applied to the different nodes and edges.

\subsection{Formalization of the Generic Algorithm}
For the complexity reasoning, we will count the different types of operations. Thus, it is convenient to formalize the algorithm as a labeled transition system
over flows and heights, the labels indicating the performed operation:
\begin{lstlisting}
  inductive_set pr_lts :: "((preflow\<times>labeling) \<times> op_ty \<times> (preflow\<times>labeling)) set" 
  where
    "push_pre f l e ==> ((f,l),PUSH,(push_eff f e,l)) \<in> pr_lts"
  | "relabel_pre f l u ==> ((f,l),RELABEL,(f,relabel_eff f l u)) \<in> pr_lts"
\end{lstlisting}  
Here, the \isai{inductive_set} construction defines the least relation that can be derived by the specified rules. 
A push operation can be implemented in time $O(1)$, (only the flow on a single edge has to be modified) and a relabel operation can be implemented in time $O(V)$.
(to find the new height, we have to consider the heights of all successor nodes in the residual graph) We define 
\isai{cost PUSH == 1 | cost RELABEL == card(V)} and prove the following theorem:
\begin{lstlisting}
theorem (in Network) generic_preflow_push_OV2E_and_correct:
  assumes "((f_0, l_0), p, (f, l)) \<in> pr_lts^*" 
  shows "(\<sum>x<-p. cost x) \<le> 38 * (card V)$^2$ * card E"
    and "(f,l)\<notin>Domain pr_lts --> isMaxFlow f"
\end{lstlisting}  
where \isai$(f_0,l_0)$ is the initial flow and height-labeling and \isai{pr_lts^*} denotes the reflexive transitive closure of \isai{pr_lts}. 
Intuitively, the cost of any (partial) run \isai$p$ from the initial state to $(f,l)$ is bounded by $O(V^2E)$, and if the 
run is complete, \ie no more operations apply (\isai$(f,l)\<notin>Domain pr_lts$), the resulting flow is maximal.

@@

The complexity argument is more tricky. Instead of defining a single measure function,
Cormen counts the different types of operations separately, using quite involved arguments.
For our formalization, we have adapted some Cormen's proofs, trading precision of the constant factors for simplicity of the formalization. 
Thus, our constant of $38$ is somewhat higher than the $17$ that would result from Cormen's proof.

@@

The simple structure of the generic algorithm allowed us to conveniently define it as labeled transition system, such that the above informal proof could be formalized
by inductive arguments over the runs of the algorithm.
Moreover, the powerful Isabelle/HOL standard library, in particular the lemmas for summations
over finite sets, were very helpful to keep the formalization effort manageable. 
Anyway, the whole complexity proof is roughly 1000 lines of Isabelle text, and took a few weeks to complete.
This includes the time that we spent on an initial termination proof that used simpler arguments, but did not derive an exact complexity bound.

\section{Instantiating the Generic Push-Relabel Algorithm}\label{sec:prpu-inst}
The generic algorithm does not specify the order in which the operations are applied. 
Choosing the right order decreases the worst-case time complexity, and, more important, the practical efficiency of the algorithm.

We consider two common instantiations of the generic algorithm: The relabel-to-front algorithm~\cite{CLRS09}, and tzhe FIFO push-relabel algorithm~\cite{GoTa88}.
Both have a theoretical time complexity of $O(V^3)$, which is better than the $O(V^2E)$ bound of the generic algorithm.
However, the more complicated structure of these algorithms prevent a convenient specification as labeled transition system, and thus 
a complexity proof in the style of the generic algorithm.
Instead, we use the Monadic Refinement Framework to specify the algorithms and only prove correctness and 
termination.\footnote{Which we get for free from the bound of the generic algorithm.} We leave it to future work to integrate complexity reasoning tools into 
the Refinement Framework and prove the tighter complexity bounds for the specific implementations.

A general idea of most efficient implementations is to \emph{discharge} a node by repeatedly applying push operations until the node gets inactive or needs to be relabeled.

\subsection{Relabel-to-Front Algorithm}
Cormen et al.\ present the relabel to front algorithm.
It discharges a node until it gets inactive, applying relabel operations as necessary. 
Moreover, if a node was discharged without being relabeled, the already saturated edges during this discharge are stored, and not considered 
during the next discharge. This reduces the work required to find the next edge to push.
The algorithm scans through a list of all nodes, discharging every node as it is visited.
If the node gets relabeled during discharging, it is moved to the front of the list, and the scan of the list begins anew.
The algorithm terminates if it reaches the end of the list.

@@

Our algorithm follows the implementation that we found in the Stanford ACM-ICPC notebook~\cite{ICPC-notebook}, 
which uses an additional gap heuristics: If there is a value $0<k<|V|$ such that there is no node that has height $k$, 
all nodes with $k<l(v)<|V|$ may be relabeled to $|V|+1$, preserving a valid labeling. 
This heuristics prevents unnecessary pushes of flow towards the sink when the path to the sink is already blocked.
The correctness of the heuristics follows easily from the definitions of a valid height-labeling.
In total, the correctness proof of the FIFO push-relabel algorithm is about 400 lines of code and, building on the experience from relabel-to-front, it 
took only a day to complete.

\section{Refinement to Executable Code}\label{sec:executable}
  In the previous section, we have presented our abstract formalization of the Edmonds-Karp and push-relabel algorithms.
  However, we left open several algorithmic aspects. For example, we did not specify how to find a shortest augmenting path.
  Moreover, some of the operations require additional auxiliary data structures to be maintained. For example, 
  we maintain the current excess of each node explicitly, instead of summing up the incoming and outgoing flows each time 
  we need to check whether a node is active.
  Finally, we have to specify the concrete data structures that we want to use for implementation. 
  For example, the residual graph is modeled as a function from edges to capacities on the abstract level. In the implementation, we use a two-dimensional array.

  In this section, we outline the further refinement steps that were necessary to obtain an efficient implementation.

>>>>

\title{Formalization of Network Flow Algorithms}
\subtitle{A Refinement Approach in Isabelle/HOL}

@@

Using stepwise refinement techniques, we instantiate the generic Ford-Fulkerson method to Edmonds-Karp algorithm, 
and the generic push-relabel algorithm of Goldberg and Tarjan to both the relabel-to-front and the FIFO push-relabel algorithm.
Further refinement then yields verified efficient implementations of the algorithms, which compare well to unverified reference implementations.

@@

\section{Introduction}
Computing the maximum flow of a network is an important problem in graph theory.
Many other problems, like maximum bipartite matching, edge disjoint paths, circulation demand, as well as various scheduling and resource allocating problems can be reduced to it.
The Ford-Fulkerson method~\cite{FF56} describes a class of algorithms to solve the maximum flow problem. 
The basic idea is to repeatedly search for \emph{augmenting paths} along which more flow can be moved from the source to the sink.
An important instance is the Edmonds-Karp algorithm~\cite{EK72},
which was one of the first algorithms to solve the maximum flow problem in polynomial time ($O(VE^2)$) for the general case of networks with real-valued capacities.
Dinitz~\cite{Di06} generalized the idea of augmenting paths to blocking flows, obtaining an $O(V^2E)$ algorithm. 
Karzanov~\cite{Ka74} was the first to propose an $O(V^3)$ algorithm based on the idea of \emph{preflows}. Based on Karzanov's ideas, Goldberg and Tarjan proposed
the generic push-relabel algorithm~\cite{GoTa88}.
Implementations of the push-relabel algorithm are among the most efficient maximum flow algorithms. While the generic algorithm has a time complexity of $O(V^2E)$,
specific variants of the algorithms achieve even lower complexities up to $O(V^2\sqrt{E})$.

In this paper, we present a formal verification of the Edmonds-Karp algorithm and of the generic push-relabel algorithm.
We prove correctness of the algorithms as well as the time complexity bounds of $O(VE^2)$ and $O(V^2E)$, respectively. 
The formalization is conducted in the Isabelle/HOL proof assistant~\cite{NPW02}. 
Stepwise refinement techniques~\cite{Wirth71,Back78,BaWr98} allow us to elegantly structure our verification to separate the abstract algorithmic ideas from
the implementation details: After proving the min-cut-max-flow theorem, it is straightforward to verify a generic version of the Ford-Fulkerson algorithm,
which we then refine to Edmonds-Karp algorithm and finally to an efficient implementation. 
Similarly, the generic push-relabel algorithm is refined to both the relabel-to-front algorithm~\cite{CLRS09} and the FIFO push-relabel algorithm~\cite{GoTa88},
which, in turn, are refined to efficient implementations. The efficiency of our verified implementations compare well to unverified reference implementations of 
the algorithms.

@@

In the rest of this paper, we give a short informal introduction the Ford-Fulkerson method (\S\ref{sec:background})
and present our formalization of it (\S\ref{sec:abs-formalization}). After introducing the Isabelle Refinement Framework~\cite{LaTu12,La12} (\S\ref{sec:refinement}), we report on the instantiation of the Ford-Fulkerson method to the Edmonds-Karp algorithm, and on its complexity proof (\S\ref{sec:edka}).
Next, we describe the correctness and complexity proof of the generic push-relabel algorithm (\S\ref{sec:gen-prpu}), and its instantiations 
to the relabel-to-front and FIFO push-relabel algorithms (\S\ref{sec:prpu-inst}).
Finally, we report on the refinement of all three algorithms to efficient implementations (\S\ref{sec:executable}), and their comparison to unverified reference 
implementations (\S\ref{sec:benchmark}).

@@

A (flow) network is a directed graph over a finite set of nodes $V$ and edges $E$, where each edge $(u,v)\in E$ is labeled by a positive real-valued capacity $c(u,v)>0$.
Moreover, there are two distinct nodes $s,t\in V$, which are called \emph{source} and \emph{sink}.

A \emph{flow} $f$ on a network is a labeling of the edges with real values satisfying the following constraints: 1) \emph{Capacity constraint}: the flow on each edge is a non-negative value smaller or equal to the edge's capacity; 2) \emph{Conservation constraint}: For all nodes except $s$ and $t$, the sum of incoming flows equals the 
sum of outgoing flows.
% 
% \emph{Non-generation constraint}: For all nodes except $s$ and $t$, the sum of flows over all incoming edges is not smaller than the sum of flows over all outgoing edges.
% We define the \emph{excess} flow $x_f(u)$ of a node $u$ to be the difference of the incoming and outgoing flow of $u$.
% Intuitively, a \emph{preflow} must not send more flow over an edge than its capacity allows, and only the source node may generate flow. 
% The excess flow is the amount of flow that is stuck at a node, \ie which flows in but does not flow out.
% A \emph{flow} is a \emph{preflow} that satisfies the \emph{Conservation constraint}: For all nodes except $s$ and $t$, the excess flow is zero.
% Intuitively, only the source may generate flow, and only the sink may consume flow, and for all other nodes, the incoming flow must be equal to the outgoing flow.
The value of a flow $f$ is denoted by $|f|$, and defined to be the sum over the outgoing flows of $s$ minus the sum over the incoming flows of $s$.
Given a network $G$, the maximum flow problem is to find a flow with a maximum value among all flows of the network.

To simplify reasoning about the maximum flow problem, we assume that our network satisfies some additional constraints: 1) the source only has outgoing edges while the sink only has incoming edges; 2) if the network contains an edge $(u, v)$ then there is no \emph{parallel edge} $(v, u)$ in the reverse direction\footnote{With $u=v$, this also implies that there are no self loops.}; and 3) every node of the network must be on a path from $s$ to $t$. Note that any network can be transformed to a network with the aforementioned properties and the same maximum flow~\cite{CLRS09}.

An important result is the relation between flows and cuts in a network. A \emph{cut} is a partitioning of the nodes into two sets, such that one set contains 
the source and the other set contains the sink. The capacity of a cut is the sum of the capacities of all edges going from the source's side to the sink's side of the cut.
It is easy to see that the value of any flow cannot exceed the capacity of any cut, as all flow from the source must ultimately reach the sink, and thus go through the edges of the cut. The min-cut-max-flow theorem tightens this bound and states that the value of the maximum flow is equal to the capacity of the minimum cut.

The Ford-Fulkerson method is a corollary of this theorem. It is based on a greedy approach: Starting from a zero flow, the value of the flow is iteratively increased until a maximum flow is reached. In order to increase the overall flow value, it may be necessary to redirect some flow, \ie to decrease the flow passed through specific edges. For this purpose the Ford-Fulkerson method defines the residual graph, which has edges in the same and opposite direction as the network edges.
% , which has edges both in parallel and reverse to the edges of the network. 
Each edge is labeled by the amount of flow that can be effectively passed along this edge, by either increasing or decreasing the flow on a network edge.
Formally, the residual graph $G_f$ of a flow $f$ is the graph induced by the 
edges with positive labels according to the following labeling function $c_f$:
\[ c_f (u, v) = 
  \begin{cases}
  c (u, v) - f(u, v) \hfill & \text{ if $(u, v) \in E$} \\
  f (v, u) \hfill & \text{ if $(v, u) \in E$} \\  
  0 \hfill & \text{ otherwise} \\
  \end{cases} 
\]

@@

The correctness of the Ford-Fulkerson algorithm follows from the min-cut-max-flow theorem, which states that the following three propositions are equivalent:
\begin{enumerate}
\item $f$ is a maximum flow in a network $G$.
\item there is no augmenting path in the residual graph $G_f$.
\item there is a cut $C$ in $G$ such that the capacity of $C$ is equal to the value of $f$.
\end{enumerate}

@@

In Isabelle/HOL, stepwise refinement is supported by the Isabelle Refinement Framework~\cite{LaTu12,La12}. 
It features a refinement calculus for programs phrased in a nondeterminism monad. 
The monad's type is a set of possible results plus an additional value that indicates a failure:
\begin{lstlisting}
  datatype \<alpha> nres = res \<alpha> set | fail
\end{lstlisting}
The operation \isai{return x} of the monad describes the single result \isai{x}, and the 
operation \isai{bind m f} nondeterministically picks a result from \isai{m} and executes \isai{f} on it. 
The bind operation fails iff either \isai{m = fail}, or \isai{f} may fail for a result in \isai{m}.

We define the \emph{refinement ordering} on \isai{\<alpha> nres} by lifting the subset ordering with \isai{fail} being the greatest element.
Intuitively, \isai{m <= m'} means that \isai{m} is a refinement of \isai{m'}, \ie all possible results of \isai{m} are 
also possible results of \isai{m'}. 
Note that the refinement ordering is a complete lattice, and bind is monotonic. Thus, we can define recursion using a fixed point construction~\cite{Kr10}.
Moreover, we can use the standard Isabelle/HOL constructs for if, let and case distinctions, yielding a fully fledged programming 
language, shallowly embedded into Isabelle/HOL's logic. For simpler usability, we define standard loop constructs (while, foreach), 
a syntax for postcondition specifications, and use a Haskell-like do-notation:
\begin{lstlisting}
  spec P == spec x. P x == res {x. P x}
  do {x <- m; f x} == bind m f
  do {m; m'} == bind m (%_. m')
\end{lstlisting}

Correctness of a program \isai{m} with precondition \isai{P} and postcondition \isai{Q} is expressed as \isai{P ==> m <= spec r. Q r} (or, eta contracted, just \isai{spec Q}), which
means that, if \isai{P} holds, \isai{m} does not fail and all possible results of \isai{m} satisfy \isai{Q}. Note that we provide different recursion constructs
for partial and total correctness: A nonterminating total correct recursion yields \isai{fail}, which satisfies no specification, even if joined 
with results from other possible runs. On the other hand, a nonterminating partial correct recursion yields \isai|res {}|, which refines any specification 
and disappears when joined with other results.

@@

Various tools, including a verification condition generator, assist the user in conducting the refinement proofs by breaking 
them down to statements that do not contain monad constructs any more. In many cases, these verification conditions precisely reflect the core idea of the proof.

@@

\section{The Generic Push-Relabel Algorithm}\label{sec:gen-prpu}
  We formalize the generic push relabel algorithm of Goldberg and Tarjan~\cite{GoTa88},
  closely following the textbook presentation of Cormen et al.~\cite{CLRS09}.

  Push-relabel algorithms are based on \emph{preflows}, a generalization of flows
  by relaxing the conservation constraint to allow more incoming flow than outgoing flow on a node.
  For a node $u\in V$, the \emph{excess} $x_f(u)$ is the difference between incoming and outgoing flow. 
  For a preflow, the excess of any node but the source must not be negative.
  A non-sink node with positive excess is called \emph{active}. Thus, a preflow without any active nodes is actually a flow.

  The push-relabel algorithm maintains a preflow $f$ and a height labeling $l: V \to \mathbb {N}$ of the nodes, 
  with the invariant that for every edge $(u, v)$ of the residual graph we have that $l(u) \le l(v) + 1$. 
  Moreover, the height of the source is fixed to $\left\vert{V}\right\vert$, and the height of the sink is fixed to $0$.
  By generalizing the labeling invariant to paths, we get that for each path $p$ from $u$ to $v$ in the residual graph, we have $l(u) \le l(v) + |p|$.
  This implies that there cannot be a path from $s$ to $t$: the length of any (simple) path is less than $|V|$, however, a valid labeling 
  implies a length greater than or equal to $|V|$. Upon termination of the algorithm, there will be no active nodes. 
  Thus, the preflow is actually a flow, and as we have no $s$--$t$ paths, the Ford-Fulkerson theorem implies that it is maximal.

  The push relabel algorithm initializes the heights of all nodes but the source to zero. Moreover, the outgoing edges of the source are 
  set to carry maximum flow, all other edges get zero flow. Then, the algorithm repeats the following operations as long as they apply:
  \begin{description}
    \item[$\textrm{push}(u,v)$] for an edge $(u,v)$ of the residual graph with $l(u) = l(v)+1$ and $u$ active: move as much flow as possible from $u$ to $v$.
    The maximum possible flow is constrained by the residual capacity of the edge, and the excess flow available at $u$.
    \item[$\textrm{relabel}(u)$] for an active node $u$ from which no push operation is possible: increase the height of $u$ to the minimum value such that a push
      becomes possible.
  \end{description}
  Obviously, if none of the operations apply, all nodes are inactive and the resulting preflow is a maximum flow.
  The algorithm is generic in the sense that it leaves open the order in which the operations are applied to the different nodes and edges.

\subsection{Formalization of the Generic Algorithm}
For the complexity reasoning, we will count the different types of operations. Thus, it is convenient to formalize the algorithm as a labeled transition system
over flows and heights, the labels indicating the performed operation:
\begin{lstlisting}
  inductive_set pr_lts :: "((preflow\<times>labeling) \<times> op_ty \<times> (preflow\<times>labeling)) set" 
  where
    "push_pre f l e ==> ((f,l),PUSH,(push_eff f e,l)) \<in> pr_lts"
  | "relabel_pre f l u ==> ((f,l),RELABEL,(f,relabel_eff f l u)) \<in> pr_lts"
\end{lstlisting}  
Here, the \isai{inductive_set} construction defines the least relation that can be derived by the specified rules. 
A push operation can be implemented in time $O(1)$, (only the flow on a single edge has to be modified) and a relabel operation can be implemented in time $O(V)$.
(to find the new height, we have to consider the heights of all successor nodes in the residual graph) We define 
\isai{cost PUSH == 1 | cost RELABEL == card(V)} and prove the following theorem:
\begin{lstlisting}
theorem (in Network) generic_preflow_push_OV2E_and_correct:
  assumes "((f_0, l_0), p, (f, l)) \<in> pr_lts^*" 
  shows "(\<sum>x<-p. cost x) \<le> 38 * (card V)$^2$ * card E"
    and "(f,l)\<notin>Domain pr_lts --> isMaxFlow f"
\end{lstlisting}  
where \isai$(f_0,l_0)$ is the initial flow and height labeling and \isai{pr_lts^*} denotes the reflexive transitive closure of \isai{pr_lts}. 
Intuitively, the cost of any (partial) run \isai$p$ from the initial state to \isai$(f,l)$ is bounded by $O(V^2E)$, and if the 
run is complete, \ie no more operations apply (\isai$(f,l)\<notin>Domain pr_lts$), then the resulting flow is maximal.

@@

The complexity argument is more tricky. Instead of defining a measure function,
Cormen counts the different types of operations separately, using quite involved arguments.
For our formalization, we have adapted Cormen's proofs, trading precision of the constant factors for simplicity of the formalization. 
Thus, our constant of $38$ is somewhat higher than the $17$ that would result from Cormen's proof.

@@

The simple structure of the generic algorithm allowed us to conveniently define it as labeled transition system, such that the above informal proof could be formalized
by inductive arguments over the runs of the algorithm.
Moreover, the powerful Isabelle/HOL standard library, in particular the support for summations
over finite sets, was very helpful to keep the formalization effort manageable. 
Anyway, the whole complexity proof is roughly 1000 lines of Isabelle text, and took a few weeks to complete.
This includes the time that we spent on an initial termination proof that used simpler arguments, but did not derive an exact complexity bound.

\section{Instantiating the Generic Push-Relabel Algorithm}\label{sec:prpu-inst}
The generic algorithm does not specify the order in which the operations are applied. 
Choosing the right order decreases the worst case time complexity, and, more important, the practical runtime.

We consider two common instantiations of the generic algorithm: The relabel-to-front algorithm~\cite{CLRS09}, and the FIFO push-relabel algorithm~\cite{GoTa88}.
Both have a theoretical time complexity of $O(V^3)$, which is better than the $O(V^2E)$ bound of the generic algorithm.
However, the more complicated structure of these algorithms prevents a convenient specification as labeled transition system, and thus 
a complexity proof in the style done for the generic algorithm.
Instead, we use the Monadic Refinement Framework to specify the algorithms and only prove correctness and 
termination.\footnote{Which we get for free from the bound of the generic algorithm.} We leave it to future work to integrate complexity reasoning tools into 
the Refinement Framework and to prove the tighter complexity bounds for the specific implementations.

A general idea of most efficient implementations is to \emph{discharge} a node by repeatedly applying push operations until the node gets inactive or needs to be relabeled.
The implementations mainly differ in the order in which the nodes are discharged.

\subsection{Relabel-to-Front Algorithm}
Cormen et al.~\cite{CLRS09} present the relabel to front algorithm.
It discharges a node until it gets inactive, applying relabel operations as necessary. 
Moreover, if a node was discharged without being relabeled, the already saturated edges during this discharge are stored, and not considered 
during the next discharge. This reduces the work required to find the next edge to push.
The algorithm scans through a list of all nodes, discharging every node as it is visited.
If the node gets relabeled during discharging, it is moved to the front of the list, and the scan of the list begins anew.
The algorithm terminates if it reaches the end of the list.

@@

Our algorithm follows the implementation that we found in the Stanford ACM-ICPC notebook~\cite{ICPC-notebook}, 
which uses an additional gap heuristics: If there is a value $0<k<|V|$ such that there is no node that has height $k$, 
all nodes $v$ with $k<l(v)<|V|$ may be relabeled to $|V|+1$, preserving a valid labeling. 
This heuristics prevents unnecessary pushes of flow towards the sink when the path to the sink is already blocked.
The correctness of the heuristics follows easily from the definitions of a valid height labeling.
In total, the correctness proof of the FIFO push-relabel algorithm is about 400 lines of code and, building on the experience from relabel-to-front, it 
took only a day to complete.

\section{Refinement to Executable Code}\label{sec:executable}
  In the previous section, we have presented our abstract formalization of the Edmonds-Karp and push-relabel algorithms.
  However, we left open several algorithmic aspects. For example, we did not specify how to find a shortest augmenting path.
  Moreover, efficient implementation of some operations requires additional auxiliary data to be maintained. 
  Finally, we have to specify the concrete data structures to be used for the implementation. 
  In this section, we outline the further refinement steps that were necessary to obtain an efficient implementation.

  \subsection{Algorithms}
  The most complex problem for which we specified no algorithm yet is finding a shortest augmenting path.
  We use breadth first search (BFS) for this. We adapted a BFS formalization from the examples collection of the Isabelle Refinement Framework to
  our needs and added an efficient imperative implementation. The resulting algorithm 
  is independent of flow networks and can be reused for other applications.
